### 2017-05-15 paper/basicSkeletons.tex
restructured some stuff. more cross references.

\subsection{Lazy \code{parEvalN}}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEvalNLazy}
	\caption{Schematic depiction of parEvalNLazy}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}. The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
\end{lstlisting} %$ %% formatting
\caption{Definition of parEvalNLazy}
\label{fig:parEvalNLazy}
\end{figure}

\subsection{Heterogenous tasks}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of parEval2}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the paralellization arrows of the same type until now. But sometimes we want to paralellize heterogenous types as well. However, we can implement such a \code{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass. The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
\\\\
We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}. Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above. After the calculation is finished, we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}. The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g = 
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{lstlisting}
	\caption{Definition of parEval2}
	\label{fig:parEval2}
\end{figure}

>>>>

\subsubsection{Lazy \code{parEvalN}}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEvalNLazy}
	\caption{Schematic depiction of parEvalNLazy}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp} (Fig.~\ref{fig:listApp}). The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
\end{lstlisting} %$ %% formatting
\caption{Definition of parEvalNLazy}
\label{fig:parEvalNLazy}
\end{figure}

\subsubsection{Heterogenous tasks}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of parEval2}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the paralellization arrows of the same type until now. But sometimes we want to paralellize heterogenous types as well. However, we can implement such a \code{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass. The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
\\\\
We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}. Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above. After the calculation is finished, we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}. The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g = 
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{lstlisting}
	\caption{Definition of parEval2}
	\label{fig:parEval2}
\end{figure}

### 2017-05-15 paper/mapSkeletons.tex
restructured some stuff. more cross references.

\subsection{Statically load-balancing parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farm}
	\caption{Schematic depiction of farm}
	\label{fig:farmImg}
\end{figure}
A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle
\end{lstlisting}
\caption{Definition of farm}
\label{fig:farm}
\end{figure}

>>>>

\subsection{Statically load-balancing parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farm}
	\caption{Schematic depiction of farm}
	\label{fig:farmImg}
\end{figure}
A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle

unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])

takeEach :: Int -> [a] -> [a]
takeEach n [] = []
takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)

shuffle :: (Arrow arr) => arr [[a]] [a]
shuffle = arr (concat . transpose)
\end{lstlisting}
\caption{Definition of farm. \code{unshuffle}, \code{takeEach}, \code{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
\label{fig:farm}
\end{figure}

### 2017-05-15 paper/parrows.tex
minor
restructured some stuff. more cross references.

\subsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{lstlisting}% $ %% formatting
\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a} (Fig.~\ref{fig:confa}). We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
data Conf a = Conf (Strategy a)
\end{lstlisting}
\caption{Definition of Conf a}
\label{fig:confa}
\end{figure}
To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} as in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{lstlisting}
\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsection{Par Monad}
The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function. The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.

We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call. This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}. In order to wait for the computation to finish, we map over the \code{IVar}s inside the ParMonad with \code{arr (>>= mapM get)}. The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{lstlisting} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsection{Eden}
For the Multicore and ParMonad implementation we could use general instances of \code{ArrowParallel} that just require the \code{ArrowApply} and \code{ArrowChoice} typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass \code{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{lstlisting}
\caption{possible ArrowUnwrap typeclass}
\label{fig:ArrowUnwrap}
\end{figure}
We don't do it here for aesthetic resons, though. For now, we just implement \code{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{lstlisting}
\caption{ArrowParallel instance for functions in the Eden backend}
\label{fig:ArrowParallelEdenFns}
\end{figure}
and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{lstlisting}
\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
\label{fig:ArrowParallelEdenKleisli}
\end{figure}

>>>>

\subsection{ArrowParallel instances}

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{lstlisting}% $ %% formatting
\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a} (Fig.~\ref{fig:confa}). We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
data Conf a = Conf (Strategy a)
\end{lstlisting}
\caption{Definition of Conf a}
\label{fig:confa}
\end{figure}
To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} as in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{lstlisting}
\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{Par Monad}
The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function. The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.

We therefore apply each function to its corresponding input value with \code{listApp} (Fig.~\ref{fig:listApp}) and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call. This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}. In order to wait for the computation to finish, we map over the \code{IVar}s inside the ParMonad with \code{arr (>>= mapM get)}. The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{lstlisting} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsubsection{Eden}
For the Multicore and ParMonad implementation we could use general instances of \code{ArrowParallel} that just require the \code{ArrowApply} and \code{ArrowChoice} typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass \code{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{lstlisting}
\caption{possible ArrowUnwrap typeclass}
\label{fig:ArrowUnwrap}
\end{figure}
We don't do it here for aesthetic resons, though. For now, we just implement \code{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{lstlisting}
\caption{ArrowParallel instance for functions in the Eden backend}
\label{fig:ArrowParallelEdenFns}
\end{figure}
and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{lstlisting}
\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
\label{fig:ArrowParallelEdenKleisli}
\end{figure}

\FloatBarrier
